---
title: "Animal rescue incidents attended by London Fire Brigade, 2009-2021"
author: "Laurence Horton"
date: "28/06/2021"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, echo = FALSE, include = FALSE}
library(tidyverse) # For data manipulation
library(leaflet) # For mapping
library(htmltools) # For popups on map
```

```{r read in data, echo = FALSE, include = FALSE}
dataset <-
  read_csv(
    "https://data.london.gov.uk/download/animal-rescue-incidents-attended-by-lfb/8a7d91c2-9aec-4bde-937a-3998f4717cd8/Animal%20Rescue%20incidents%20attended%20by%20LFB%20from%20Jan%202009.csv",
    col_names = FALSE
  ) # Import without column names because it cases an error.
dataset[1, 8] <-
  "HourlyNotionalCost" # remove £ pound sign from vector
dataset[1, 9] <-
  "IncidentNotionalCost" # remove £ pound sign from vector
names(dataset) <-
  dataset %>% slice(1) %>% unlist() # move first row to column names
dataset <- dataset %>% slice(-1)
```

```{r reformat borough as upper case, echo = FALSE, include = FALSE}
dataset$Borough <-
  toupper(dataset$Borough) # Change borough names to upper case for consistency
```

## Animal rescue incidents attended London Fire Brigade by London borough, 2009-2021
I wondered what London borough had the greatest number of incidents over the period covered by the data set.
```{r visualisation by borough, echo = FALSE}
borough <- dataset %>%
  select(Borough) %>%
  group_by(Borough) %>%
  summarise(Incidents = n()) %>% # Count number of incidents by borough
  filter(Incidents > 1, na.rm = TRUE) %>% # Remove boroughs with less than one incident
  na.omit() # Remove incidents without a borough

# Visualisation
ggplot(data = borough, aes(x = reorder(Borough, (Incidents)), y = Incidents)) +
  geom_bar(stat = "identity") +
  labs(
    y = "Incidents",
    x = "",
    caption = "Data: London Fire Brigade",
    title = "Animal rescue incidents attended by London Fire Brigade",
    subtitle = "By London borough, 2009-2021"
  ) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_line(colour = "grey", size = 0.25),
    panel.grid.minor.x = element_blank(),
    panel.background = element_rect(fill = "#FFFFFF", color = "white"),
  ) +
  scale_fill_brewer(palette = "Greys") +
  coord_flip()
```

## Animal rescue incidents attended London Fire Brigade by region, 2009-2021
London has an inner and outer region, so what's the pattern of incidents for those regions?
```{r visualistation by region and year, echo = FALSE, message = FALSE}
# Recode boroughs into London region, either Inner or Outer London.
region <- dataset %>%
  select(CalYear, Borough) %>%
  na.omit() %>%
  mutate(
    Region = case_when(
      Borough == "CAMDEN" |
        Borough == "GREENWICH" |
        Borough == "HACKNEY" |
        Borough == "HAMMERSMITH AND FULHAM" |
        Borough == "ISLINGTON" |
        Borough == "KENSINGTON AND CHELSEA" |
        Borough == "LAMBETH" |
        Borough == "LEWISHAM" |
        Borough == "SOUTHWARK" |
        Borough == "TOWER HAMLETS" |
        Borough == "WANDSWORTH" | Borough == "WESTMINSTER" ~ "Inner",
      Borough != "CAMDEN" |
        Borough != "GREENWICH" |
        Borough != "HACKNEY" |
        Borough != "HAMMERSMITH AND FULHAM" |
        Borough != "ISLINGTON" |
        Borough != "KENSINGTON AND CHELSEA" |
        Borough != "LAMBETH" |
        Borough != "LEWISHAM" |
        Borough != "SOUTHWARK" |
        Borough != "TOWER HAMLETS" |
        Borough != "WANDSWORTH" | Borough != "WESTMINSTER" ~ "Outer"
    )
  ) %>%
  group_by_at(vars(Region, CalYear)) %>%
  summarise(Incidents = n())

# Manipulate subset for visualisation
inner <- region %>%
  select(Region, CalYear, Incidents) %>%
  filter(Region == "Inner") %>%
  rename(Inner = Incidents)

outer <- region %>%
  select(Region, CalYear, Incidents) %>%
  filter(Region == "Outer") %>%
  rename(Outer = Incidents)

# Calculate percentages for region by year.
region_year <- inner_join(inner, outer, by = "CalYear")
region_year <- region_year %>%
  mutate(
    InnerPct = round(Inner / (Inner + Outer) * 100, digits = 1),
    OuterPct = round(Outer / (Inner + Outer) * 100, digits = 1)
  )

# Split them for a merge
inner_pct <- region_year %>%
  select(c(Region.x, CalYear, InnerPct)) %>%
  rename(Region = Region.x, Pct = InnerPct, Year = CalYear)

outer_pct <- region_year %>%
  select(c(Region.y, CalYear, OuterPct)) %>%
  rename(Region = Region.y, Pct = OuterPct, Year = CalYear)

# Merge
region_year_pct <- rbind(inner_pct, outer_pct)

# Remove tables no longer needed.
rm(inner, outer)

# Visualisation of % share of incidents by year and region.
ggplot(data = region_year_pct, aes(x = Year, y = Pct, fill = Region)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(Pct, sep = "%")),
            position = position_stack(vjust = .5),
            size = 2.5) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(
    y = "%",
    x = "Year",
    caption = "Data: London Fire Brigade",
    title = "Animal rescue incidents attended by London Fire Brigade by region",
    subtitle = "2009-2021"
  ) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_line(colour = "grey", size = 0.25),
    panel.grid.minor.x = element_blank(),
    panel.background = element_rect(fill = "#FFFFFF", color = "white"),
  ) +
  scale_fill_brewer(palette = "Greys") +
  coord_flip()
```

## Map of animal rescue incidents attended London Fire Brigade in Canning Town, 2009-2021
I used to live in Canning Town so I wondered what kind of animal rescue incidents had occurred there.
```{r map animal rescue incidents for Canning Town, echo = FALSE}
# Create geobox for Canning Town
n <- (51.5177) # north
s <- (51.5070) # south
e <- (0.0064) # east
w <- (0.0347) # west

# Create map
map <-
  dataset %>%
  drop_na() %>% # Drop cases with no geocordinates
  select(SpecialServiceType, Latitude, Longitude) %>%
  filter(Latitude < n &
           Latitude > s &
           Longitude > e &
           Longitude < w)
rm(n, s, e, w)

# Convert longitude and latitude to numeric
map$Latitude <- as.numeric(map$Latitude)
map$Longitude <- as.numeric(map$Longitude)

# Add to map
leaflet(map) %>% addTiles() %>%
  addMarkers( ~ Longitude, ~ Latitude, popup = ~ htmlEscape(SpecialServiceType))
```

# Citations
## Data
London Fire Brigade (2021) Animal Rescue incidents attended by LFB from Jan 2009. *London Datastore* https://data.london.gov.uk/dataset/animal-rescue-incidents-attended-by-lfb available under an UK Government Open Licence (OGL v3) http://www.nationalarchives.gov.uk/doc/open-government-licence/version/3/

## Code
Horton, L. (2021). #TidyTuesday 2021-06-29 LFB Animal Rescues. GitHub repository, https://github.com/laurencehorton/tidytuesday
published under Creative Commons Zero v1.0 Universal https://github.com/laurencehorton/torontobikeshare/blob/master/LICENSE

## Packages
Mock, T. (2021). *Tidy Tuesday: A weekly data project aimed at the R ecosystem*. https://github.com/rfordatascience/tidytuesday.

Cheng, J., Karambelkar, B., Xie, Y. (2019). *leaflet: Create Interactive Web Maps with the JavaScript 'Leaflet' Library*. R package version 2.0.3. https://CRAN.R-project.org/package=leaflet

Cheng, J., Sievert, C., Chang, W., Xie, Y., Allen, J. (2021). *htmltools: Tools for HTML*. R package version 0.5.1.1. https://CRAN.R-project.org/package=htmltools

Mock, T. (2021). *Tidy Tuesday: A weekly data project aimed at the R ecosystem*. https://github.com/rfordatascience/tidytuesday.

R Core Team (2018). *R: A language and environment for statistical computing. R Foundation for Statistical Computing*, Vienna, Austria. URL https://www.R-project.org/.

Wickham, H., François, R., Henry, L., Müller K. (2021). *dplyr: A Grammar of Data Manipulation. R package version 1.0.6*. https://CRAN.R-project.org/package=dplyr

```{r citations, echo = FALSE, include = FALSE}
citation()
citation("dplyr")
citation("leaflet")
citation("htmltools")
sessionInfo()
```

